The Driving Forces of API Evolution

### 2015-06-01 thesis/paper.tex
Toned-down findings & clarified API/APIs

\begin{abstract}
%Motivation
Evolving an Application Programming Interface (API) is a precipitous activity, as modifications to them can significantly impact their users. The increasing use of APIs means that software development organisations must take an empirical and scientific approach to the way they manage the evolution of their APIs. 
%Problem
If no attempt at analysing or quantifying the evolution of an API is made, there will be a diminished understanding of the evolution, and possible improvements to the maintenance strategy will be difficult to identify. We believe that long-standing software evolution theories can provide additional insight to the field of APIs, and can be of great use to companies maintaining APIs. 
%Method
In this case study, we conduct a qualitative investigation to understand what drives the evolution of an industry company's existing API, by examining two versions of the API code. The changes were analysed based on two software evolution theories, and the extent to which we could reverse engineer the change decisions was determined by interviewing the architect of the API. 
%Results
The results of this analysis show that the largest driving force of API evolution is the desire for new functionality. Our findings which show that changes happen sporadically, rather than continuously, invalidate the law of Conservation of Organisational Stability in the context of APIs. We also found that it is possible to reverse engineer change decisions and in doing so, identified that the feedback loop of an API is the most important area of improvement. 
\end{abstract}

@@

The study was conducted in two major phases \cite{andersson2007spiral}. The aim of the first phase was to generate hypotheses and to formulate clear research questions. This was achieved by using a data-driven approach to analysing the code and documentation. The second phase included analysis of the interviews, during which we aimed to confirm the hypotheses that were formulated in the first phase. Both phases were performed iteratively. This created the opportunity to improve the data analysis as the study progressed, as well as, allowing us to adapt to possible changes in direction, due to the hypothesis generating approach of the first phase. An overview of the methodology can be found in Table \ref{table:meth_steps}.

@@

Our results show that the major driving force for change in APIs is the call for increased functionality. The classification of the types of software evolution and software maintenance showed that enhancive changes which either aimed to add modules or new functionality to existing modules were the most common. We also think that it is interesting to note that a majority of the changes which did not affect functionality, such as the updative changes, were spurred because of previous changes that were related to functionality. Following this line of thought, the desire for new functionality can be seen as a catalyst that allows new change to occur. The view of considering functionality to be a major driver for change is supported by the analysis with Lehman's laws. In the law of Continuing Change, Lehman concludes that ``an E-type system must be continually adapted or it becomes progressively less satisfactory". The results of our work supports this view, as the changes did not only add functionality, but it was also found that the application developers most major concern were updates to functionality.

@@

The results also showed that usability was the second-most important driver of API evolution. The classified updative, groomative and reformative changes all affected the maintainability and usability of the API and involved activities that have been found to improve the usability and maintainability of an API \cite{afonso2012evaluating, clarke2004measuring, piccioni2013empirical, shi2011empirical}. The interviewee discussed how company A wanted to aid developers creating applications by improving the interfaces that the developers are using. This improvement to the interfaces would better communicate the design and intentions to the developer. Improving the documentation of the API was also important for v2.0, as were improvements to the example code. These updates were intended to aid developers and create a set of standards that should be followed when creating applications. These changes correspond with three of Lehman's laws: Conservation of Familiarity, Declining Quality and Feedback Loop.

Our results found that seven out of eight of Lehman's laws do apply to API evolution. The law of Conservation of Organisational Stability was found not to apply, since changes to APIs are unfavourable due to the potential disruption the applications using the API could face. This inertia to change in API evolution is supported by other research, which also conclude that the negative effect that changes to APIs may have is a major factor to be considered \cite{google_talk, henning2007api, mcdonnell2013empirical, robbes2012developers}.

Based on our results, we form the conclusion that the desire for functionality and the reluctance of changing are two opposing forces that cause updates to APIs to be infrequent. In the evolution of APIs, the negative effects of changes are greater than in other types of systems, which is the main cause for the law of Conservation of Organisational Stability not applying to APIs. This occurring has been considered by Lehman, as he noted that ``the laws are not immutable, since they arise from the habits and practices of humans" \cite{lehman1980programs}. The underlying reason for this is that the laws are influenced by conscious thought processes, based on human understanding \cite{lehman1980understanding}, which in the case of APIs is the conscious decision to limit updates as much as possible.

@@

By analysing the API based on Lehman's laws, we were able to find that the law of Conservation of Organisational Stability did not apply to the API examined in this study. Our results suggest that the opposite of what the law suggests, in its current form, is more applicable for APIs. Since only a single API was studied, we cannot safely refute the law or propose a new formulation of it. Research should therefore be wary of assuming that this law is true for APIs.

@@

\section{Conclusion} \label{conclusion}
With this study, we conclude that studying API evolution with help of existing theories, provides a new and structured approach to the field of API evolution. Our study shows that functionality is the largest driving force of change in the evolution of APIs. It also concludes that APIs do not change invariantly, and that maintenance of an API rather is infrequent and sporadic. This finding lead us to invalidate Lehman's law of Conservation of Organisational Stability in the context of APIs. With this study we have also identified the importance of having a structured feedback loop when maintaining an API. We have also provided company A with a foundation for facilitating such a change.

@@

To further investigate what drives API evolution, we believe that API evolution has to be studied from a business point of view. We therefore suggest that a similar study is conducted, but with the goal of analysing the change of an API based on business motives. Positioning a study around business motives would help identifying the changes before they are formulated into technical changes, and it would provide a new and fresh angle to the field of API evolution.

>>>>

\begin{abstract}
%Motivation
Evolving an Application Programming Interface (API) is a precipitous activity, as modifications to them can significantly impact their users. The increasing use of APIs means that software development organisations must take an empirical and scientific approach to the way they manage the evolution of their APIs. 
%Problem
If no attempt at analysing or quantifying the evolution of an API is made, there will be a diminished understanding of the evolution, and possible improvements to the maintenance strategy will be difficult to identify. We believe that long-standing software evolution theories can provide additional insight to the field of APIs, and can be of great use to companies maintaining APIs. 
%Method
In this case study, we conduct a qualitative investigation to understand what drives the evolution of an industry company's existing API, by examining two versions of the API code. The changes were analysed based on two software evolution theories, and the extent to which we could reverse engineer the change decisions was determined by interviewing the architect of the API. 
%Results
The results of this analysis show that the largest driving force of the APIs evolution was the desire for new functionality. Our findings which show that changes happen sporadically, rather than continuously, contradict the law of Conservation of Organisational Stability in the context of APIs. We also found that it is possible to reverse engineer change decisions and in doing so, identified that the feedback loop of an API is an important area of improvement. 
\end{abstract}

@@

The study was conducted in two phases. The aim of the first phase was to generate hypotheses and to formulate clear research questions. This was achieved by using a data-driven approach to analysing the code and documentation. The second phase included analysis of the interviews, during which we aimed to confirm the hypotheses that were formulated in the first phase. Both phases were performed iteratively. This created the opportunity to improve the data analysis as the study progressed, as well as, allowing us to adapt to possible changes in direction, due to the hypothesis generating approach of the first phase. An overview of the methodology can be found in Table \ref{table:meth_steps}.

@@

Our results show that the major driving force for change the case company's API is the call for increased functionality. The classification of the types of software evolution and software maintenance showed that enhancive changes which either aimed to add modules or new functionality to existing modules were the most common. We also think that it is interesting to note that a majority of the changes which did not affect functionality, such as the updative changes, were spurred because of previous changes that were related to functionality. Following this line of thought, the desire for new functionality can be seen as a catalyst that allows new change to occur. The view of considering functionality to be a major driver for change is supported by the analysis with Lehman's laws. In the law of Continuing Change, Lehman concludes that ``an E-type system must be continually adapted or it becomes progressively less satisfactory". The results of our work supports this view, as the changes did not only add functionality, but it was also found that the application developers most major concern were updates to functionality.

@@

The results also showed that usability was the second-most important driver of the studied API's evolution. The classified updative, groomative and reformative changes all affected the maintainability and usability of the API and involved activities that have been found to improve the usability and maintainability of an API \cite{afonso2012evaluating, clarke2004measuring, piccioni2013empirical, shi2011empirical}. The interviewee discussed how company A wanted to aid developers creating applications by improving the interfaces that the developers are using. This improvement to the interfaces would better communicate the design and intentions to the developer. Improving the documentation of the API was also important for v2.0, as were improvements to the example code. These updates were intended to aid developers and create a set of standards that should be followed when creating applications. These changes correspond with three of Lehman's laws: Conservation of Familiarity, Declining Quality and Feedback Loop.

Our results found that seven out of eight of Lehman's laws do apply to the evolution of the studied API. The law of Conservation of Organisational Stability was found not to apply, since changes to APIs are unfavourable due to the potential disruption the applications using the API could face. This inertia to change in API evolution is supported by other research, which also conclude that the negative effect that changes to APIs may have is a major factor to be considered \cite{google_talk, henning2007api, mcdonnell2013empirical, robbes2012developers}.

Based on our results, we form the conclusion that the desire for functionality and the reluctance of changing are two opposing forces that cause updates to this API to be infrequent. In the evolution of APIs, the negative effects of changes are greater than in other types of systems, which is the main indication that the law of Conservation of Organisational Stability is not applicable to APIs. This occurrence has been considered by Lehman, as he noted that ``the laws are not immutable, since they arise from the habits and practices of humans" \cite{lehman1980programs}. The underlying reason for this is that the laws are influenced by conscious thought processes, based on human understanding \cite{lehman1980understanding}, which in the case of APIs is the conscious decision to limit updates as much as possible.

@@

By analysing the API based on Lehman's laws, we were able to find that the law of Conservation of Organisational Stability did not apply to the API examined in this study. Our results suggest that the opposite of what the law suggests, in its current form, is more applicable for APIs. However, since only a single API was studied, we cannot generalise this finding for all APIs. Research should therefore be wary of assuming that this law is true for APIs and future work could be undertaken to fully determine it's validity in regards to APIs.

@@

\section{Conclusion} \label{conclusion}
With this study, we conclude that studying API evolution with help of existing theories, provides a new and structured approach to the field of API evolution. Our study shows that functionality is the largest driving force of change in the evolution of a single API. It also concludes that this API did not change invariantly, and that maintenance of an its maintenance was infrequent and sporadic which appears to contradict one of Lehman's law; the law of Conservation of Organisational Stability. However, given that this study was on performed on a single API, we cannot generalise this finding. Furthermore, we have identified the importance of having a structured feedback loop when maintaining an API. We have also provided company A with a foundation for facilitating such a change.

@@

To further investigate what drives API evolution, we believe that API evolution has to be studied from a business point of view. We therefore suggest that a similar study is conducted, but with the goal of analysing the change of an API based on business motives. Positioning a study around business motives would help with identifying the changes before they are formulated into technical changes, and it would provide a new and fresh angle to the field of API evolution.
